Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2022-07-22 paper/ch_potentials.tex
further text improvements

% TODO rewrite, make self contained
The paper introduced the basic CH-Potentials heuristic, some query optimizations for handling of low-degree vertices, common in road networks, and some applications of CH-Potentials.
In this article, we discuss and evaluate Lazy RPHAST, the central ingredient of CH-Potentials, in greater detail and introduce an improved variant for Customizable Contraction Hierarchies.
% Furthermore, we evaluate Lazy RPHAST in the context of one-to-many and many-to-many problems.
The original paper only discussed unidirectional CH-Potentials.
In this journal version, we also discuss a bidirectional variant and introduce an improved pruning criterion for symmetrical bidirectional A*.
The improved pruning is not specific to CH-Potentials and can also be used with other heuristics.
Compared to the conference paper, the journal version contains a significantly expanded experimental evaluation.

The rest of the article is organized as follows.
In Section~\ref{sec:related_work}, we discuss related works on goal directed search and extensions for realistic applications for hierarchical techniques.
After establishing preliminaries in Section~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and Lazy RPHAST, our algorithm to solve it efficiently, in Section~\ref{sec:lazy-rphast}.
CH-Potentials, our new A* heuristic is introduced in Section~\ref{sec:main-algo}.
There, we also discuss our improvements to the handling of low-degree vertices (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
% Section~\ref{sec:extensions} presents a multitude of applications of CH-Potentials.
In Section~\ref{sec:extensions}, we demonstrate flexibility of our approach, by describing how to apply the approach to different practical applications.
Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.

@@

@@

In this work, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem (Section~\ref{sec:lazy-rphast}).
In Section~\ref{sec:main-algo}, we present CH-Potentials, a novel fast and tight heuristic for A* in road networks build on Lazy RPHAST.
Further, we present several useful optimizations for A* in road networks accelerating the processing of low degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
In Section~\ref{TODO}, we introduce a unified framework to apply CH-Potentials to a variety of practical route planning problems without any modifications to the preprocessing.
Finally, an extensive experimental evaluation in Section~\ref{sec:experiments} confirms the applicability of our results.

@@

Given vertices $s$ and $t$ we want to obtain a $st$-path $P=(s=v_0,\dots,t=v_k)$ of minimum weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.
If there is no path, we set $\dist_w(s,t)=\infty$. % TODO edge notation, degree (undirected)

@@

Dijkstra's algorithm can also be run from the target $t$ on the reversed graph $\overleftarrow{G}$. % TODO introduce reverse dist etc
We call this a \emph{backward search}.
Running two Dijkstra searches simultaneously, one from $s$ and a backward search from $t$, until the searches meet is called \emph{bidirectional} search.
In this case, we denote by $\overrightarrow{D}$, $\overrightarrow{Q}$ and $\overrightarrow{k}$ the distances, queue and minimum queue key of the forward search and by $\overleftarrow{D}$, $\overleftarrow{Q}$ and $\overleftarrow{k}$ the respective data of the backward search.
Typically, the searches are interleaved by alternating settling a vertex from each direction.
Another common approach is to always settle a vertex from the direction with the smaller minimum queue key.
Theoretically, any direction interleaving strategy is possible.
When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.

@@

\subsection{Contraction Hierarchies (CH)}

@@

\draw[dashed] (4) -- (6);
\draw[dashed] (6) -- (8);
\end{tikzpicture}
\caption{
Solid lines are edges in $G$. Dotted lines are shortcuts. Red is a shortest $st$-path in $G$. Blue is equally long up-down $st$-path in $G^+$. $v^*$ is the mid vertex. % TODO different letter for m
}
\label{fig:ch}
\Description[Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.]{
Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.
The y-position on the figure indicates the rank of each vertex.
Where the original shortest path down and then up, a shortcut edge connects the two higher vertices.
The path in the augmented graph goes only up and then down.
}
\end{figure}

A CH is a two-phase technique to efficiently compute exact, shortest paths on road networks.
For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give an introduction.

\begin{algorithm2e}[tb]
\KwData{$D^{\downarrow}[v]$: tentative distance from any vertex $v \in V$ to target $t$ in $\gchd$}
\KwData{Minimum priority queue $Q$, ordered by tentative distances}
$D^{\downarrow}[v] \gets +\infty$ for all $v\neq t$\;
$D^{\downarrow}[t] \gets 0$\;
Make $Q$ only contain $t$ with tentative distance $0$\;
\While{$Q$ not empty}{
  $u\gets$ pop minimum element from $Q$\;
  \For{all down-edges $vu \in \echd$}{
    \If{$D^{\downarrow}[v] > w(vu) + D^{\downarrow}[u]$}{
      $D^{\downarrow}[v]\gets w(vu) + D^{\downarrow}[u]$\;
      Add $v$ or decrease $v$'s key in $Q$ to $D^{\downarrow}[v]$\;
    }
  }
}
\caption{CH backward search.}
\label{algo:ch-backward}
\end{algorithm2e}

In a CH, each vertex gets a unique rank representing its ``importance''.
By convention, important vertices are usually depicted at the ``top'' while unimportant vertices are positioned at the ``bottom''.
The intuition is that dead-ends are unimportant while highways or bridges are important.
An edge goes \emph{up} when its tail vertex has lower rank than its head vertex.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one vertex $v^*$ is more important than both its neighbors.
$v^*$ is called the \emph{mid} vertex.
An \emph{up path} is a path where the last vertex is the mid vertex.
Similarly, for a \emph{down path}, the first vertex has to be the mid vertex.
% Every up and down path is an up-down path.
%
In the preprocessing phase, first, a vertex ordering is computed heuristically.
Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting the least important vertex and adding shortcuts between its higher ranked neighbors.
%See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of vertices $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for an illustration.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
By construction, the mid vertex must be in the intersection of the forward and backward search spaces.
Pseudocode for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
% TODO \gchu and \gchd

A CH query is fast, if the number of vertices reachable via only up- or down-edges is small.
On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.

% Using the CH query algorithm, we can already give a simple heuristic.
% The heuristic evaluation $h_t(x)$ performs a CH-query from $x$ to $t$.
% This yields tight estimates but a high overhead for the heuristic evaluation.
% While a single CH query is fast, answering one for every vertex explored in the $A^*$ search is slow.
% Fortunately, we can do better.
% For this we need another component called PHAST~\cite{dgnw-phast-13}.

\subsection{Customizable Contraction Hierarchies (CCH)}
\label{sec:cch-intro}

Some hierarchical techniques were extended to support a three-phase setup.
The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
The preprocessing phase typically takes hours, the customization seconds, and the query milliseconds.
The preprocessing phase is independent of the weights and only exploits the graph topology.
Weights are introduced in the customization phase.
Paths are computed in the query phase.
This setup is useful to allow for weight changes that are common to all vehicles such as live traffic updates.
The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}. % TODO better clear up whats related work and whats prelim

Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that follows this three-phase setup.
The key idea is to use a vertex importance order that is independent of the weight function.
CCH use \emph{nested dissection} orders for this, i.e.\ a small balanced vertex separator is computed and the separator vertices get the highest importance.
The remaining vertices are ordered by recursing on the independent cells obtained by removing the separator vertices.
This results in surprisingly good CH orders.
Regardless of the weight function, any shortest path between the cells has to use some of the separator vertices, therefore these vertices will always be ``important''.

Such orders can be obtained with partitioning algorithms tailored to road networks~\cite{GottesburenHUW19} in less than an hour even for continental-sized road networks.
Once the order was computed, the augmented graph $G^+$ is computed.
The standard CH preprocessing algorithms could be used for this but specialized, significantly faster CCH variants exist.
However, discussing them is beyond the scope of this work.
See~\cite{dsw-cch-15,BuchholdSW19} for details.
A CCH augmented graph fulfills all necessary properties for CH query algorithms.
The CH query algorithms can therefore be applied without modifications.
However, the CCH preprocessing grants some useful additional properties.

During preprocessing, an \emph{elimination tree} can be constructed.
The root of the elimination tree is the most important vertex.
For every other vertex its parent is its least important upward neighbor.
In~\cite{bcrw-s-16}, it was proven that the set of ancestors of a vertex in the elimination tree is equal to the vertices reachable in a CH search.
This makes it possible to explore the CCH search space with a more efficient shortest path algorithm:
For the forward search, start at the source and follow the elimination tree upward until the root is reached.
All outgoing edges of encountered vertices are relaxed.
The backward search works analogously but starts at the target.
Because this elimination tree-based algorithm does not require a queue, it is faster than a Dijkstra-based CCH query.
With this algorithm, CCH can consistently answer shortest path queries on continental sized road networks in fractions of a milliseconds.
CH queries are still marginally faster for weight functions with a strong hierarchy such as travel times.
However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly and CCH queries are the faster variant.
Note that only nested dissection orderings admit an elimination tree of low height.
The elimination tree query can therefore not be used for a general CH.

@@

\section{CH-Potentials: A New A* Heuristic}\label{sec:main-algo}

In this section, we describe how to use the Lazy RPHAST algorithm to build a fast and tight heuristic for A*.
We also discuss CH-Potentials for the bidirectional A* algorithm and propose an improved pruning criterion for symmetric bidirectional potentials.
Further, we present several low degree vertex optimizations which exploit road network characteristics to reduce the overhead of queue operations and heuristic evaluations.
Finally, we introduce an algorithmic framework to apply A* with CH-Potentials to a variety of practical route planning problems.

The \emph{CH-Potentials} heuristic is a straightforward application of the Lazy RPHAST algorithm.
In the beginning of each query, we perform the target selection, a backward CH search, from the target $t$.
The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $v$.

This yields a \emph{perfect} heuristic, at least as long as the CH preprocessing and the A* algorithm are performed on the same graph with the same weight function.
This case is of course not particularly interesting.
One could just answer the shortest path query directly with a CH query.
This approach becomes useful when the query runs on a \emph{different but related} graph or weight function than the preprocessing.
We will discuss such applications in Section~\ref{sec:extensions}.
But before that, we discuss two extensions/optimizations to A* which we will utilize.

@@

\subsection{Applications}\label{sec:extensions}

% A* is a flexible algorithm.
Many extended problems can be solved using CH-Potentials.
We start by establishing a common formal framework for the use of CH-Potentials.
Then, we exemplary describe some extended routing problems and how to apply CH-Potentials to them.

\subsubsection{Formal Setup: Inputs, Outputs, and Phases}

% TODO make sure to distinguish PROBLEM model from SOLUTION appraoch

We consider a variety of different of applications, with slightly different problem models.
The goal is always to quickly answer many shortest path queries.
For the purpose of describing our framework, we establish a shared notation:
Input to each query are vertices $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
In the simplest case, $w_q$ will be scalar edge weights.
However, this is not a requirement.
It can be any function that computes a weight for an edge.
This function can also take additional parameters from the state of the search.
For example, in the case of live-traffic, $w_q$ represents scalar edge weights.
However, values of $w_q$ might change between queries.
In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.

To enable quick shortest path computations, we consider a two-phase setup with an additional offline preprocessing phase before the online query phase.
The input to the preprocessing phase is a graph $G_\ell = (V_{\ell}, E_{\ell})$ with lower bound weights $w_\ell$.
$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
The output of the preprocessing is auxiliary data that allows to quickly compute distances on $G_\ell$ with respect to $w_\ell$.
In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.

The query phase uses this auxiliary data in an A* search between vertices $s$ and $t$ on $G_q = (V_q, E_q)$ with weights $w_q$.
Let $\phi : V_q \to V_{\ell}$ denote a function mapping vertices in the query graph to vertices in the lower bound graph.
We require that $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$ with respect to $w_\ell$.
Then, A* uses the heuristic function $h_t(x) = \dist_{w_{\ell}}(\phi(x), \phi(t))$ computed using Lazy RPHAST.
Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph, $\phi$ is the identity function, and only $w_q$ changes for the queries.
The exact implementation of the A* search depends on the application.
Our approach only provides the heuristic $h_t$ for the A* search.
In contrast, the preprocessing phase remains the same for all applications.

Our heuristic is always \emph{feasible}, i.e. $w_q(u v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
By requirement and because of the triangle inequality the following holds:
\[
w_q(u v) - h_t(u) + h_t(v) \geq \dist_\ell(\phi(u), \phi(v)) - \dist_\ell(\phi(u), \phi(t)) + \dist_\ell(\phi(v), \phi(t)) \geq 0
\]
Thus, A* will always determine the correct shortest distances.

\subsubsection{Avoiding Tunnels and/or Highways}
\label{sec:no-tunnel-highway}

@@

The classical shortest path problem allows to freely change edges at vertices.
However, in the real world, turn restrictions, such as a forbidden left or right turn, exist.
Also, taking a left turn might take longer than going straight.
This can be modeled using turn weights~\cite{gv-errnt-11,dgpw-crprn-13,bwzz-cchtc-20}.
% In this section, we first present the extended problem setting.
% Afterwards, we describe how it can be solved with CH-Potentials.
A \emph{turn weight} $w_t : E \times E \to \mathbb{R}^{\geq 0}$ maps a pair of incident edges onto the turning time or $+\infty$ for forbidden turns. % TODO formally
For CH-Potentials, we use zero as lower bound for every turn weight in the heuristic.
Thus, the graph $G_\ell$ and weights $w_\ell$ for preprocessing is the unmodified input graph without turn weights.

@@

If all travel time functions fulfill the \emph{FIFO property}, this problem can be solved using a straightforward extension of Dijkstra's algorithm \cite{d-aassp-69}.
The necessary modification to A* is analogous.
Without the FIFO property the problem becomes NP-hard \cite{or-tnp-89}.
% The FIFO property states that it is never beneficial to wait at a vertex before entering an edge.
The FIFO property states that it is not possible to arrive earlier by departing later.
Formally stated, the following must hold $\forall e\in E,\tau\in \mathbb{R},\delta\in \mathbb{R}^+: w_q(e, \tau) \le w_q(e, \tau+\delta) + \delta$.
Our implementation stores edge travel times using piece-wise linear functions.
The A* search uses the tentative distance $\tau$ at a vertex $u$ when evaluating the travel time of outgoing edges $u v$.
This strategy is very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.

For the preprocessing, we set $w_\ell(e) = \min_\tau w_q(e,\tau)$, that is the minimum travel time.
% With this setup, we extended CH-Potentials to support time-dependent routing.
By keeping travel time functions out of the CH, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-21,dn-crdtd-12} which have to create shortcuts of travel time functions.

@@

\subsubsection{CCH Potentials}

Live traffic can alternatively be supported using a three-phase setup as described in Section~\ref{sec:cch-intro}.
We do this by exchanging the CH for a CCH.
The preprocessing phase is identical as for a CCH and is independent of $w_\ell$.
The $w_\ell$ is fixed during the customization phase.
It can be changed arbitrarily by rerunning the customization.
The A* algorithm is part of the query phase.
The weights $w_q$ used in the query phase must not be smaller than the $w_\ell$ used during the last customization.

This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic}.
Changes to the live traffic situation are handled by changing $w_\ell$ via a customization.
Additional features such as turn restrictions can be implemented by adjusting $w_q$.
For severe traffic events, the three-phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
However, it comes at the expense of a more complex algorithm.
Further, when integrating predicted traffic, CCH-Potentials will not have much advantage over CH-Potentials because the heuristic must be a lower bound for the entire time considered and not only the current situation.

@@

% \subsubsection{Alternative Routes with the Penalty Method}

>>>>

@@

% classical
% goal, hierarchy, labeling
% extended query (m2o)
% extended application

% TODO RPHAST etc

@@

After discussing notation and fundamental algorithms in Section~\ref{sec:preliminaries}, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem in Section~\ref{sec:lazy-rphast}.
Section~\ref{sec:astar_opts} contains several optimizations for A* in road networks accelerating the processing of low degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
Our main contribution, the CH-Potentials framework is presented in Section~\ref{sec:framework} where we show how to use Lazy RPHAST to build a fast and tight heuristic for A* in road networks.
Based on this heuristic, we provide a unified framework (Section~\ref{sec:formal_framework} and~\ref{sec:chpot}) for a variety of practical route planning problems (Section~\ref{sec:applications}).
CH-Potentials can be applied to all of these problems without any modifications to the preprocessing.
In Section~\ref{sec:experiments}, we provide an extensive experimental evaluation analyzing the performance characteristics of our algorithms.
It shows that CH-Potentials achieve decent running times when tight lower bounds are available at preprocessing time.
However, the strength of our approach is not in query running times but in its flexibility:
Problem extensions can be supported at the cost of some slowdown to the queries but without any algorithmic adjustments.

@@

Given vertices $s$ and $t$ we want to obtain a $st$-path $P=(s=v_0,\dots,t=v_k)$ of minimum weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.
If there is no path, we set $\dist_w(s,t)=\infty$.

@@

Dijkstra's algorithm can also be run from the target $t$ on the reversed graph $\overleftarrow{G}$.
We call this a \emph{backward search}.
Running two Dijkstra searches simultaneously, one from $s$ and a backward search from $t$, until the searches meet is called \emph{bidirectional} search.
In this case, we denote by $\overrightarrow{D}$, $\overrightarrow{Q}$ and $\overrightarrow{k}$ the distances, queue and minimum queue key of the forward search and by $\overleftarrow{D}$, $\overleftarrow{Q}$ and $\overleftarrow{k}$ the respective data of the backward search.
Typically, the searches are interleaved by alternating settling a vertex from each direction.
Another common approach is to always settle a vertex from the direction with the smaller minimum queue key.
Theoretically, any direction interleaving strategy is possible.
When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.

@@

\subsection{Contraction Hierarchies}

@@

\draw[dashed] (4) -- (6);
\draw[dashed] (6) -- (8);
\end{tikzpicture}
\caption{
Solid lines are edges in $G$. Dotted lines are shortcuts. Red is a shortest $st$-path in $G$. Blue is equally long up-down $st$-path in $G^+$.
}
\label{fig:ch}
\Description[Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.]{
Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.
The y-position on the figure indicates the rank of each vertex.
Where the original shortest path down and then up, a shortcut edge connects the two higher vertices.
The path in the augmented graph goes only up and then down.
}
\end{figure}

\begin{algorithm2e}[tb]
\KwData{$D^{\downarrow}[v]$: tentative distance from any vertex $v \in V$ to target $t$ in $\gchd$}
\KwData{Minimum priority queue $Q$, ordered by tentative distances}
$D^{\downarrow}[v] \gets +\infty$ for all $v\neq t$\;
$D^{\downarrow}[t] \gets 0$\;
Make $Q$ only contain $t$ with tentative distance $0$\;
\While{$Q$ not empty}{
  $u\gets$ pop minimum element from $Q$\;
  \For{all reversed downward edges $uv \in \rechd$}{
    \If{$D^{\downarrow}[u] + \overleftarrow{w}(uv) < D^{\downarrow}[v]$}{
      $D^{\downarrow}[v] \gets D^{\downarrow}[u] + \overleftarrow{w}(uv)$\;
      Add $v$ or decrease $v$'s key in $Q$ to $D^{\downarrow}[v]$\;
    }
  }
}
\caption{CH backward search.}
\label{algo:ch-backward}
\end{algorithm2e}

\emph{Contraction Hierarchies} (CH) is a two-phase speed-up technique to accelerate shortest path computations on road networks through precomputation.
For a detailed discussion we refer to~\cite{gssv-erlrn-12}.
Here, we only briefly introduce necessary notation and algorithms used in this paper.
In a preprocessing phase, vertices are ordered totally by ``importance'' where more important vertices should lie on more shortest paths.
Intuitively, vertices on highways are more important than vertices on rural streets.
For CH, such an ordering is obtained heuristically.
The position of a vertex in the order is also denoted as its \emph{rank}.
Vertices of higher rank are informally referred to and visualized as ``higher up'' in the hierarchy.
Therefore, an edge where the tail has lower rank than the head is an \emph{upward} edge.
Analogously, when the head vertex has the lower rank, the edge is said to go \emph{downward}.
Once such an importance ordering was obtained, all vertices are contracted successively by ascending importance.
To \emph{contract} a vertex means to temporarily remove it from the graph while inserting \emph{shortcut edges} between more important neighbors to preserve shortest distances among them.
The result is an \emph{augmented graph} $G^+$ with original edges and shortcuts.
We often refer to the augmented graph split into an upward graph $\gchu = (V, \echu)$ containing only upward edges and a downward graph $\gchd = (V, \echu)$ containing only downward edges.

The augmented graph has the property that for any two vertices $s$ and $t$, there always exists an \emph{up-down} $st$-path, i.e.\ a path  which first uses only edges from $\gchu$ and then only edges from $\gchd$, with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for an illustration.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
Such an up-down path can be found by running Dijkstra's algorithm from $s$ on $\gchu$ and from $t$ on the reversed downward graph $\rgchd$ graph.
By construction, at least the highest ranked vertex on the up-down path must be in the intersection of the forward and backward search spaces.
Pseudocode for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
The set of vertices reachable in $\gchu$ and $\rgchd$ is called the \emph{CH search space} of a vertex.
The CH query performance strongly correlates with the size of this CH search space and the number of edges between the vertices in the search space.
Luckily, on road networks, the CH search space is small~\cite{gssv-erlrn-12,dgpw-crprn-13}.

\subsection{Customizable Contraction Hierarchies}
\label{sec:cch-intro}

Some hierarchical techniques were extended to support a three-phase setup~\cite{dgpw-crprn-13}.
The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
The preprocessing phase typically takes hours, the customization seconds, and the query milliseconds.
The preprocessing phase is independent of the weights and only uses the graph topology.
Weights are introduced in the customization phase.
Paths are computed in the query phase.

Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) is a variant of Contraction Hierarchies that follows this three-phase setup.
The key idea is to use a vertex importance order that is independent of the weight function.
CCH use \emph{nested dissection} orders for this.
A small balanced vertex separator is computed and the separator vertices get the highest importance.
The remaining vertices are ordered by recursing on the independent cells obtained by removing the separator vertices.
This results in surprisingly good CH orders.
Regardless of the weight function, any shortest path between the cells has to use some of the separator vertices, therefore these vertices will always be ``important''.
Such orders can be obtained with partitioning algorithms tailored to road networks~\cite{GottesburenHUW19} in less than an hour even for continental-sized road networks.
Once the order was computed, the augmented graph $G^+$ is computed.
The standard CH preprocessing algorithms could be used for this but specialized, significantly faster CCH variants exist.
However, discussing them in detail is beyond the scope of this work.
See~\cite{dsw-cch-15,BuchholdSW19} for details.
A CCH augmented graph fulfills all necessary properties for CH query algorithms.
The CH query algorithms can therefore be applied without modifications.
However, the CCH preprocessing grants some useful additional properties.

During preprocessing, an \emph{elimination tree} can be constructed.
The root of the elimination tree is the most important vertex.
For every other vertex its parent is its least important upward neighbor.
In~\cite{bcrw-s-16}, it was proven that the set of ancestors of a vertex in the elimination tree is equal to its CH search space.
This makes it possible to explore the CCH search space with a more efficient shortest path algorithm:
For the forward search, start at the source and follow the elimination tree upward until the root is reached.
All outgoing edges of encountered vertices are relaxed.
The backward search works analogously but starts at the target.
Because this elimination tree-based algorithm does not require a queue, it is faster than a Dijkstra-based CCH query.
With this algorithm, CCH can consistently answer shortest path queries on continental sized road networks in fractions of a millisecond.
CH queries are still marginally faster for weight functions with a strong hierarchy such as travel times.
However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly and CCH queries turn out the faster variant.
Note that only nested dissection orderings admit an elimination tree of low height.
The elimination tree query can therefore not be used with a classical CH.

@@

\section{Optimizations for A* in Road Networks}\label{sec:astar_opts}

In this section, we propose optimizations for A* in road networks.
First, we present several low degree vertex optimizations which exploit road network characteristics to reduce the overhead of queue operations and heuristic evaluations.
Second, we discuss the bidirectional A* algorithm and propose an improved pruning criterion for symmetric bidirectional potentials.
These optimizations can be used with \emph{any} A* heuristic.

@@

\section{The CH-Potentials Framework}\label{sec:framework}

% , we describe how to use the Lazy RPHAST algorithm to build a fast and tight heuristic for A*.
In this section, we introduce an algorithmic framework to apply A* with a Lazy RPHAST-based heuristic to a variety of practical route planning problems.
We call our approach \emph{CH-Potentials}.
The core idea is to compute a CH augmented graph during preprocessing and use A* with a straightforward application of Lazy RPHAST as the heuristic to answer queries.
However, when the CH preprocessing and the A* algorithm are performed on the same graph with the same weight function, this yields a \emph{perfect} heuristic
This case is, of course, not particularly interesting.
One could just answer the shortest path query directly with a CH query.
The approach becomes useful when the query runs on a \emph{different but related} graph or weight function than the preprocessing.
Therefore, we start by establishing a common formal framework for the use of CH-Potentials.
Then, we exemplarily describe some extended routing problems and how to apply the CH-Potentials framework to them.

\subsection{Formal Setup: Inputs, Outputs, and Phases}\label{sec:formal_framework}

We consider a variety of different of applications, with slightly different problem models.
The goal is always to quickly answer many shortest path queries.
For the purpose of describing our framework, we establish a shared notation:
Input to each query are vertices $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
In the simplest case, $w_q$ will be scalar edge weights.
Live traffic is an example for this.
The challenge in this scenario is that values of $w_q$ might change between queries.
However, $w_q$ can also represent something more complex than scalar numbers.
It can be any function that computes a weight for an edge.
This function can also take additional parameters from the state of the search.
In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.

To enable quick shortest path computations, we consider a two-phase setup with an additional offline preprocessing phase before the online query phase.
The input to the preprocessing phase is a lower bound graph $G_\ell = (V_{\ell}, E_{\ell})$ with lower bound weights $w_\ell$ where $w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
The output of the preprocessing is auxiliary data that allows to quickly compute distances on $G_\ell$ with respect to $w_\ell$.
In the applications considered in this paper, $w_\ell$ is always the \emph{freeflow} travel time.

The query phase may use this auxiliary data to answer shortest path queries between vertices $s$ and $t$ on $G_q = (V_q, E_q)$ with weights $w_q$.
Let $\phi : V_q \to V_{\ell}$ denote a function mapping vertices in the query graph to vertices in the lower bound graph.
The crucial requirement for a routing problem to fit into our framework is that the query weight of an edge $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ between the corresponding vertices $\phi(u)$ and $\phi(v)$ in the lower bound graph $G_\ell$ with respect to $w_\ell$.
Typically, and unless stated otherwise, $G_q$ and $G_\ell$ are the same graph, $\phi$ is the identity function, and only $w_q$ changes for the queries.

\subsection{CH-Potentials}\label{sec:chpot}
CH-Potentials can be used to solve any problem in this framework.
The preprocessing is always the computation of the CH augmented graph and remains the same regardless of the specific routing problem.
The query consists of A* with the heuristic function $h_t(v) = \dist_{w_{\ell}}(\phi(v), \phi(t))$ computed using Lazy RPHAST.
In the beginning of each query, we perform the target selection, i.e.\ a backward CH search, from the target $t$.
The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $v$ (see Algorithm~\ref{algo:pot}).
Note that, in contrast to the preprocessing, the exact implementation of the A* search depends on the application.
Our approach only provides the heuristic $h_t$ for the A* search.
As the performance of A* depends crucially on the accuracy of the heuristic estimates, the smaller the difference between query weights and lower bound distances, the better CH-Potentials will perform.

\paragraph{Correctness}
Our heuristic is always \emph{feasible}, i.e. $w_q(u v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
By requirement and because of the triangle inequality the following holds:
\[
w_q(u v) - h_t(u) + h_t(v) \geq \dist_\ell(\phi(u), \phi(v)) - \dist_\ell(\phi(u), \phi(t)) + \dist_\ell(\phi(v), \phi(t)) \geq 0
\]
Thus, A* will always determine the correct shortest distances.

\subsection{Applications}\label{sec:applications}
\subsubsection{Avoiding Tunnels and/or Highways}
\label{sec:no-tunnel-highway}

@@

The classical shortest path problem allows to freely change edges at vertices.
However, in the real world, turn restrictions, such as a forbidden left or right turn, exist.
Also, taking a left turn might take longer than going straight.
This can be modeled using turn weights~\cite{gv-errnt-11,dgpw-crprn-13,bwzz-cchtc-20}.
% In this section, we first present the extended problem setting.
% Afterwards, we describe how it can be solved with CH-Potentials.
A \emph{turn weight} $w_t : E \times E \to \mathbb{R}^{\geq 0}$ maps a pair of incident edges onto the turning time or $+\infty$ for forbidden turns.
For CH-Potentials, we use zero as lower bound for every turn weight in the heuristic.
Thus, the graph $G_\ell$ and weights $w_\ell$ for preprocessing is the unmodified input graph without turn weights.

@@

If all travel time functions fulfill the \emph{FIFO property}, this problem can be solved using a straightforward extension of Dijkstra's algorithm \cite{d-aassp-69}.
The necessary modification to A* is analogous.
Without the FIFO property the problem becomes NP-hard \cite{or-tnp-89}.
% The FIFO property states that it is never beneficial to wait at a vertex before entering an edge.
The FIFO property states that it is not possible to arrive earlier by departing later.
Formally stated, the following must hold $\forall e\in E,\tau\in \mathbb{R},\varepsilon\in \mathbb{R}^{>0}: w_q(e, \tau) \le w_q(e, \tau+\varepsilon) + \varepsilon$.
Our implementation stores edge travel times using piecewise linear functions.
The A* search uses the tentative distance $\tau$ at a vertex $u$ when evaluating the travel time of outgoing edges $u v$.
This strategy is very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.

For the preprocessing, we set $w_\ell(e) = \min_\tau w_q(e,\tau)$, that is the minimum travel time.
% With this setup, we extended CH-Potentials to support time-dependent routing.
By keeping travel time functions out of the CH preprocessing, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-21,dn-crdtd-12} which have to create shortcuts of travel time functions.

@@

\subsubsection{Live Traffic with CCH-Potentials}

Live traffic can alternatively be supported using a three-phase setup as described in Section~\ref{sec:cch-intro}.
For this, we use CCH instead of CH preprocessing to build the augmented graph.
This allows us to introduce updates to $w_\ell$ and the augmented graph within a few seconds using the CCH customization algorithm.
This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic} where live traffic was represented only in $w_q$.
In contrast, with CCH-Potentials, changes to the live traffic situation are handled by also changing $w_\ell$ via a customization.
Additional features such as turn restrictions can be implemented by adjusting $w_q$.
For severe traffic events, the three-phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
However, it comes at the expense of a more complex setup.
Also, the customization is not fast enough to rerun it for every query.
Further, when integrating predicted traffic, CCH-Potentials will not have much advantage over CH-Potentials because the heuristic must be a lower bound for the entire time considered and not only the current situation.

@@

% \subsection{Alternative Routes with the Penalty Method}
